---
layout: page
title: "Lesson Title"
author: Adapted from XX by XX
output: html_document
---

```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Define the directory where images generated by knit will be saved
knitr::opts_chunk$set(fig.path = "images/")

# Set the web address where R will look for files from this repository
# Do not change this address
#repo_url <- "https://raw.githubusercontent.com/jescoyle/R-course-NEON/master/"
```

```{r, echo=FALSE, purl=FALSE, message = FALSE}
# This initiates styles for knitr needed by all lessons.
# TODO: MODIFY setup.R script to contain the above options and include data specifically for this NEON tutorial
#source("setup.R")
```

## Overview
### What we will learn

Brief description of the contents of this lesson including objectives:

- Objective 1
- Objective 2

### Motivator

Describe a problem that learners will be able to solve by the end of this lesson.

### Get set up

Describe any data sets or packages that learners need to download or install before proceeding.

```{r results = 'hide'}
# Read in a data file from the data folder included with the student workbook
dat <- read.csv("data/gapminder_data.csv")
```

Make a code block that loads any packages needed for the analysis. It can be helpful to indicate which functions come from which packages.

```{r results = 'hide', warning = FALSE, message = FALSE}

# Load dplyr: filter
library(dplyr)

# Load ggplot2: ggplot
library(ggplot2)

```

## Lesson


```{r}
# This is an R code block that will execute any code inside and show the results.
dat %>%
  filter(country == "United States")

```



```{r usa_pop_fig, echo = FALSE}
# echo = FALSE means that the results of the code will be shown without the code.
# chl_fig is a name that knit will use to name the figure generated by this code chunk

# Plot year vs. population size for United Stes
dat %>%
  filter(country == "United States") %>%
  ggplot(aes(x = year, y = pop)) +
  geom_line() +
  labs(x = "Year", y = "Population size")

```

**Figure 1. Population growth in the United States 1952 - 2007.** 

### Section 1


> ***TRY THIS:*** Include in-line exercises using callouts.

### Section 2


## Exercises

Number all exercises using the module number. Include starter code in the /exercises folder if appropriate.

0.1 Exercise 1 for module 0.

0.2 Exercise 2 for module 0.



## Further Reading

