---
title:  
output:
  html_document:
    toc: false
bibliography: references.bib
---


```{r, include = FALSE, purl = FALSE}
# Functions and data
getwd()
source("src/setup.R")

# get chapter number based on name
.chp <- "intro-to-R"
.chp_num <- .get_chpNum("chp-intro-to-R", type = "fileName")
.chp_str <- ifelse(nchar(.chp_num) > 1, .chp_num, paste0("0", .chp_num))

# knitr options
source("src/knit-options.R")
.knitr_fig_path(paste0(chp_str, "-"))

# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

# Chapter `r .chp_num`. `r get_chpName(.chp_num, type.to = "nameLong")` {.tabset}

## Overview 

> MOTIVATOR

Brief description of the contents of this lesson. 

### What we will learn



In this lesson 
`r .get_LO(as.numeric(.chp_num), .LOtable, prefix = TRUE)`

#### Key Terms & Commands

* term 1
* term 2
* `function1()`


### Prerequisites

Describe any data sets or packages that learners need to download or install before proceeding.

```{r results = 'hide'}
# Read in a data file from the data folder included with the student workbook
dat <- read.csv("data/gapminder_data.csv")
```


## Lesson


```{r}
# This is an R code block that will execute any code inside and show the results.
5+3

```


### Section 1


::: {.challenge}
What should be the output for each of these commands?

```{r eval = FALSE}
log(exp(3 + 3)) 
e-1 + 5
```

<details><summary>**Solution:**</summary>

::: {.solution}

```{r echo = FALSE}
log(exp(3 + 3)) 
2e-1 + 5
```

:::

</details>

:::

<br>

### Section 2

Callouts are formatted in the style.css file.

::: {.callout-trick}

:::


::: {.callout-tip}

:::

::: {.callout-important}

:::

## Exercises

After completing these exercises, learners will be able to 
`r .get_LO(.chp, .CStable, prefix = TRUE, bullet = "1.")`


Number all exercises using the module number. Include starter code in the /exercises folder if appropriate.

**`r .chp_num`.1** 

**`r .chp_num`.2** 



## Sources and Resources

This lesson was adapted from XX by Jes Coyle.

### Additional Resources

### Cited References

::: {#refs}
:::

