---
title:  
output:
  html_document:
    toc: false
bibliography: references.bib
---


```{r, include = FALSE, purl = FALSE}
# Functions and data
getwd()
source("src/setup.R")

# get chapter number based on name
chp_num <- get_chpNum("chp-computer-navigation", type = "fileName")
chp_str <- ifelse(nchar(chp_num) > 1, chp_num, paste0("0", chp_num))

# knitr options
source("src/knit-options.R")
knitr_fig_path(paste0(chp_str, "-"))

# Silently load in the data for this lesson
gapminder <- read.csv("data/gapminder_data.csv", header = TRUE)

# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

# Chapter `r chp_num`. `r get_chpName(chp_num, type.to = "nameLong")` {.tabset}

## Overview 

### What we will learn

Brief description of the contents of this lesson. 

In this lesson 
`r get_LO(as.numeric(chp_num), LOtable, prefix = TRUE)`

### Motivator

Describe a problem that learners will be able to solve by the end of this lesson.

### Get set up

Describe any data sets or packages that learners need to download or install before proceeding.

```{r results = 'hide'}
# Read in a data file from the data folder included with the student workbook
dat <- read.csv("data/gapminder_data.csv")
```

Make a code block that loads any packages needed for the analysis. It can be helpful to indicate which functions come from which packages.

```{r results = 'hide', warning = FALSE, message = FALSE}

# Load dplyr: filter
library(dplyr)

# Load ggplot2: ggplot
library(ggplot2)

```

## Lesson


```{r}
# This is an R code block that will execute any code inside and show the results.
dat %>%
  filter(country == "United States")

```



```{r usa_pop_fig}
#| echo = FALSE,
#| fig.cap = paste0("Fig. ", chp_num, ".1. Population growth in the United States 1952 - 2007.")

# Note that figure caption numbers are not automatic here.

# Plot year vs. population size for United States
dat %>%
  filter(country == "United States") %>%
  ggplot(aes(x = year, y = pop)) +
  geom_line() +
  labs(x = "Year", y = "Population size")

```


### Section 1


::: {.challenge}
> **Try this:** What should be the ouput for each of these commands?
> ```{r eval = FALSE}
> log(exp(3 + 3)) 
> 2e-1 + 5
> ```
:::
::: {.solution}
> ```{r echo = FALSE}
> log(exp(3 + 3)) 
> 2e-1 + 5
> ```
:::```

### Section 2


## Exercises

After completing these exercises, learners will be able to 
`r get_LO(as.numeric(chp_num), CStable, prefix = TRUE, bullet = "1.")`


Number all exercises using the module number. Include starter code in the /exercises folder if appropriate.

`r chp_num`.1 

`r chp_num`.2 



## Sources and Resources

This lesson was adapted from XX by Jes Coyle.

### Additional Resources

### Cited References


<!-- javascript that formats solution blocks clickable hidden text -->

```{js, file = "solution_format.java"}

```
