---
title:  
output:
  html_document:
    code_folding: show
    toc: false
bibliography: references.bib
---


```{r, include = FALSE, purl = FALSE}
# Functions and data
source("src/setup.R")

# get chapter number based on name
chp_num <- get_chpNum("chp-computer-navigation", type = "fileName")
chp_str <- ifelse(nchar(chp_num) > 1, chp_num, paste0("0", chp_num))

# knitr options
source("src/knit-options.R")
knitr_fig_path(paste0(chp_str, "-"))

# Silently load in the data for this lesson
gapminder <- read.csv("data/gapminder_data.csv", header = TRUE)

# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

# Chapter `r chp_num`. `r get_chpName(chp_num, type.to = "nameLong")` {.tabset}

## Overview 

*from:* [@swc-reproducible-lesson](https://swcarpentry.github.io/r-novice-gapminder/) episodes 4, 5, 6, 11

### What we will learn

Brief description of the contents of this lesson.


- Basic data types in R
  - vectors vs. lists
  - multidimensional arrays
  - dataframes vs. matrices
  - data types: numeric, character, factor, logical
- Manipulating data frames
- Subsetting vectors and data frames


In this lesson 
`r get_LO(as.numeric(chp_num), LOtable, prefix = TRUE)`

### Motivator

Describe a problem that learners will be able to solve by the end of this lesson.

### Get set up

Describe any data sets or packages that learners need to download or install before proceeding.

```{r results = 'hide'}
# Read in a data file from the data folder included with the student workbook
dat <- read.csv("data/gapminder_data.csv")
```

Make a code block that loads any packages needed for the analysis. It can be helpful to indicate which functions come from which packages.

```{r results = 'hide', warning = FALSE, message = FALSE}

# Load dplyr: filter
library(dplyr)

# Load ggplot2: ggplot
library(ggplot2)

```

## Lesson



### Comparing things

We can also do comparisons in R:

```{r}
1 == 1  # equality (note two equals signs, read as "is equal to")
```

```{r}
1 != 2  # inequality (read as "is not equal to")
```

```{r}
1 < 2  # less than
```

```{r}
1 <= 1  # less than or equal to
```

```{r}
1 > 0  # greater than
```

```{r}
1 >= -9 # greater than or equal to
```

> ## Tip: Comparing Numbers
>
> A word of warning about comparing numbers: you should
> never use `==` to compare two numbers unless they are
> integers (a data type which can specifically represent
> only whole numbers).
>
> Computers may only represent decimal numbers with a
> certain degree of precision, so two numbers which look
> the same when printed out by R, may actually have
> different underlying representations and therefore be
> different by a small margin of error (called Machine
> numeric tolerance).
>
> Instead you should use the `all.equal` function.
>
> Further reading: [http://floating-point-gui.de/](http://floating-point-gui.de/)
>
{: .callout}




```{r}
# This is an R code block that will execute any code inside and show the results.
dat %>%
  filter(country == "United States")

```



```{r usa_pop_fig}
#| echo = FALSE,
#| fig.cap = paste0("Fig. ", chp_num, ".1. Population growth in the United States 1952 - 2007.")

# Note that figure caption numbers are not automatic here.

# Plot year vs. population size for United States
dat %>%
  filter(country == "United States") %>%
  ggplot(aes(x = year, y = pop)) +
  geom_line() +
  labs(x = "Year", y = "Population size")

```


### Section 1


> ***CHALLENGE:*** What will this code do?
>
> ```r
> 1 + 1
> ```
> > ```{r class.source = 'fold-hide', collapse = TRUE}
> > 1 + 1
> > ```

### Section 2


## Exercises

After completing these exercises, learners will be able to 
`r get_LO(as.numeric(chp_num), CStable, prefix = TRUE, bullet = "1.")`


Number all exercises using the module number. Include starter code in the /exercises folder if appropriate.

`r chp_num`.1 

`r chp_num`.2 



## Sources and Resources

This lesson was adapted from [@swc-reproducible-lesson](https://swcarpentry.github.io/r-novice-gapminder/) episodes 1 by Jes Coyle.

### Additional Resources

### Cited References
