---
title:  
output:
  html_document:
    toc: false
bibliography: references.bib
---


```{r, include = FALSE, purl = FALSE}
# Functions and data
getwd()
source("src/setup.R")

# get chapter number based on name
.chp <- "ggplot2-part2"
.chp_num <- .get_chpNum("ggplot2-part2", type = "fileName")
.chp_str <- ifelse(nchar(.chp_num) > 1, .chp_num, paste0("0", .chp_num))

# knitr options
source("src/knit-options.R")
.knitr_fig_path(paste0(.chp_str, "-"))

# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

# Chapter `r .chp_num`. `r .get_chpName(.chp_num, type.to = "nameLong")` {.tabset}

## Overview 

> MOTIVATOR

Brief description of the contents of this lesson. 

### What we will learn

In this lesson 
`r .get_LO(.chp, .LOtable, prefix = TRUE)`

#### Key Terms & Commands

* `round()`
* `format()`


### Prerequisites

Before beginning this lesson you should have completed the lesson [`r .get_chpName("ggplot2-part1", type.from = "chp")`](chp-ggplot2-part1.html) and any prerequisites therein.

Before you begin:

1. Open the R-course-NEON-workbook RStudio project.
2. Create a new R script named `r paste0("lesson_", .chp_str, "_code.R")` in the lesson-code folder of the student workbook.
3. Whenever you see R code like this:

```{r eval = FALSE}
Type (or copy) this into your R script. Then run it in the Console.

```

4. Whenever you see a challenge like this:

:::{.challenge}

Try to solve the problem by writing R code into your script. Check your answer by clicking:

<details><summary>**Solution:**</summary>

::: {.solution}

```{r eval = FALSE}
The code that solves the challenge will appear here.
```

Along with an explanation.

:::

</details>

:::

<br>


4. Be sure to click **Save** often to save your work.

```{r include = FALSE}
# Load ggplot2
library("ggplot2")

# Set theme options
# ggplot theme options
theme_set(theme_gray(base_size = 30))
update_geom_defaults("point", aes(size = 3))


# Read in a data file from the data folder included with the student workbook

```


## Lesson

As in lesson [`r .get_chpName("program-flow", type.from = "chp")`](chp-program-flow.html), we are going to download and work with surface data  from four NEON aquatic sites: Caribou Creek, Alaska (CARI), Martha Creek, Washington (MART), Teakettle Creek, California (TECR) and Sycamore Creek, Arizona (SYCA). You can see a map of these sites on the [NEON website](https://www.neonscience.org/field-sites/explore-field-sites). However in this lesson we will be working with chemical properties of plankton  ([[DP1.20053.001](https://data.neonscience.org/data-products/DP1.20053.001) ](https://data.neonscience.org/data-products/DP1.20053.001)). 

We will download data from NEON using the `loadByProduct()` function in the `neonUtilities` package. We need to specify the product ID number, sites and dates for which we want data and then type `y` when prompted to download the data at the command line.

```{r eval = FALSE}
library(neonUtilities)

# Define dates for download: June 1 - Oct 31, 2021
# but we need to start before June 1 to get the entire month of June
startdate <- "2021-05-31"
enddate <- "2021-10-31"

# Define the sites to download 
# TECR = Teakettle Creek, SYCA = Sycamore Creek
# MART = Martha Creek, CARI = Caribou Creek
sites <- c("TECR", "SYCA", "MART", "CARI") 

# Define the data product
dpID <- "DP1.20053.001"  # Surface water temperature

# Temperature in surface water
watertemp <- loadByProduct(dpID = dpID,
                           site = sites,
                           startdate = startdate,
                           enddate = enddate)

```
```{r include = FALSE, message = FALSE}
# Load neonUtilities package
library(neonUtilities)

# Load previously downloaded object
#save(watertemp, file = "src/chp-program-flow_watertemp.RData")
load("src/chp-program-flow_watertemp.RData")
```

You should now have a new object named `watertemp` in the Environment panel. It should be a list with nine elements. Let's extract the list element of `watertemp` that contains the data and save it as a separate data frame.

```{r}
# Extract 30 min temperature averages from watertemp
# and save as a separate data frame
TSW_30min <- watertemp$TSW_30min

```

Now we have a data frame where each row summarizes temperature measurements over a 30 minute period from May - October, 2021 for the four sites from which we requested data (TECR, SYCA, MART, CARI).

The last thing we will do to get these data ready to work with is to convert the `startDateTime` column to a date-time class so that we can use it as an actual date. We will use the `ymd_hms()` function from the `lubridate()` package for this.

```{r message = FALSE}
# Load the lubridate package
library(lubridate)

# Convert startDateTime to a date-time class
TSW_30min$startDateTime <- ymd_hms(TSW_30min$startDateTime)

```

### Adding layers


add horizontal and vertical reference lines to a plot

add multiple data layers to a plot and change the order in which they are layered	


#### Statistical summaries

add a trendline to a scatterplot using statistics

add error bars to a bar graph

### Manipulating the axes



#### Position scales

describe how position scales control the way axes are displayed in ggplot2
"change the scale, limits and breaks on an axis"

#### Axis labels


### Manipulating the legend

#### Text labels

change the default labels in a legend (using factor levels or other method)

#### Color palettes for data

select inclusive color palettes that are visible to people with impaired  color vision
change a discrete and continuous color scale

### Multi-panel plots

display data in separate panels of a graph using facets


### Themes


describe how themes control the way plots are displayed in ggplot2
"change the default fonts, colors, and sizes using a theme











::: {.challenge}


<details><summary>**Solution:**</summary>

::: {.solution}



:::

</details>

:::

<br>

### Section 2

Callouts are formatted in the style.css file.

::: {.callout-trick}

:::


::: {.callout-tip}

:::

::: {.callout-important}

:::

## Exercises

After completing these exercises, learners will be able to 
`r .get_LO(.chp, .CStable, prefix = TRUE, bullet = "1.")`


ggplot2-part2	1	display data in separate panels of a graph using facets	use the ggplot2 package in R to
ggplot2-part2	2	add multiple data layers to a plot and change the order in which they are layered	use the ggplot2 package in R to
ggplot2-part2	3	add a trendline to a scatterplot using statistics	use the ggplot2 package in R to
ggplot2-part2	4	add error bars to a bar graph	use the ggplot2 package in R to
ggplot2-part2	5	"change the scale, limits and breaks on an axis"	use the ggplot2 package in R to
ggplot2-part2	6	"change the default fonts, colors, and sizes using a theme"	use the ggplot2 package in R to
ggplot2-part2	7	change a discrete and continuous color scale	use the ggplot2 package in R to
ggplot2-part2	8	change the text labels in a legend	use the ggplot2 package in R to
ggplot2-part2	9	add horizontal and vertical reference lines to a plot	use the ggplot2 package in R to



**`r .chp_num`.1** 

**`r .chp_num`.2** 



## Sources and Resources

This lesson was written by Jes Coyle and adapted from  [@swc-reproducible-lesson](https://swcarpentry.github.io/r-novice-gapminder/) episode 8.

### Additional Resources

### Cited References

::: {#refs}
:::

