---
title:  
output:
  html_document:
    code_folding: show
    toc: false
bibliography: references.bib
---


```{r, include = FALSE, purl = FALSE}
# Functions and data
source("src/setup.R")

# get chapter number based on name
chp_num <- get_chpNum("chp-functions", type = "fileName")
chp_str <- ifelse(nchar(chp_num) > 1, chp_num, paste0("0", chp_num))

# knitr options
source("src/knit-options.R")
knitr_fig_path(paste0(chp_str, "-"))

# Silently load in the data for this lesson
gapminder <- read.csv("data/gapminder_data.csv", header = TRUE)

# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```

# Chapter `r chp_num`. `r get_chpName(chp_num, type.to = "nameLong")` {.tabset}


**THIS LESSON IS NOT YET READY FOR LEARNERS.**

## Overview 

*from:* [@swc-reproducible-lesson](https://swcarpentry.github.io/r-novice-gapminder/) episodes 9, 10


### What we will learn

Brief description of the contents of this lesson. 



- NEON Utilities functions & reading a function help file

In this lesson 
`r get_LO(as.numeric(chp_num), LOtable, prefix = TRUE)`

### Motivator

Describe a problem that learners will be able to solve by the end of this lesson.

### Get set up

Describe any data sets or packages that learners need to download or install before proceeding.

```{r results = 'hide'}
# Read in a data file from the data folder included with the student workbook
dat <- read.csv("data/gapminder_data.csv")
```

Make a code block that loads any packages needed for the analysis. It can be helpful to indicate which functions come from which packages.

```{r results = 'hide', warning = FALSE, message = FALSE}

# Load dplyr: filter
library(dplyr)

# Load ggplot2: ggplot
library(ggplot2)

```

## Lesson


```{r}
# This is an R code block that will execute any code inside and show the results.
dat %>%
  filter(country == "United States")

```



```{r usa_pop_fig}
#| echo = FALSE,
#| fig.cap = paste0("Fig. ", chp_num, ".1. Population growth in the United States 1952 - 2007.")

# Note that figure caption numbers are not automatic here.

# Plot year vs. population size for United States
dat %>%
  filter(country == "United States") %>%
  ggplot(aes(x = year, y = pop)) +
  geom_line() +
  labs(x = "Year", y = "Population size")

```


### Section 1


> ***CHALLENGE:*** What will this code do?
>
> ```r
> 1 + 1
> ```
> > ```{r class.source = 'fold-hide', collapse = TRUE}
> > 1 + 1
> > ```

### Section 2

#### What are programs?

Let's write a small program

- Programs
  - programs are files written in executable code
  - we will learn to write small "programs" called "scripts"
    - you can write programs to automate tasks
    - you can write programs to make your scientific analyses reproducible

Find the R program on your computer

#### Separate function definition and application

One of the more effective ways to work with R is to start by writing the code you want to run directly in a .R script, and then running the selected lines (either using the keyboard shortcuts in RStudio or clicking the "Run" button) in the interactive R console.

When your project is in its early stages, the initial .R script file usually contains many lines
of directly executed code. As it matures, reusable chunks get pulled into their
own functions. It's a good idea to separate these functions into two separate folders; one
to store useful functions that you'll reuse across analyses and projects, and
one to store the analysis scripts.

## Exercises

After completing these exercises, learners will be able to 
`r get_LO(as.numeric(chp_num), CStable, prefix = TRUE, bullet = "1.")`


Number all exercises using the module number. Include starter code in the /exercises folder if appropriate.

`r chp_num`.1 

`r chp_num`.2 



## Sources and Resources

This lesson was adapted from [@swc-reproducible-lesson](https://swcarpentry.github.io/r-novice-gapminder/) episode 2 and XXXXXXX by Jes Coyle.

### Additional Resources

### Cited References
